---
title: "Common dplyr idioms for the life sciences"
author: "Sean Hughes / Claire Levy"
date: "Monday, January 26, 2015"
output: html_document
---

## Introduction

Life scientists often need to perform several common tasks with data generated by instruments: 

* average replicate rows
* normalize experimental rows to control rows
* find and exclude outliers

The R package [dplyr](http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html) package makes data manipulation easy. The idioms described below use `dplyr` to perform these common tasks. 

We'll use a data set from a simple experiment to illustrate these idioms. `expt` contains data from an experiment where samples from four different sources (A, B, C, and D) were exposed to DrugX or DrugY at concentrations of 0, 1, 10, 100, or 1000. Every condition was performed in duplicate. 

```{r message=FALSE, warning=FALSE, results='hide'}
library(dplyr)
expt <- read.csv("expt.csv")
head(expt)
#>     Sample Treatment Concentration Replicate      Data
#> 1 Sample A    Drug X             0         1 57.000000
#> 2 Sample A    Drug X             0         2 22.000000
#> 3 Sample A    Drug X             1         1 33.000000
#> 4 Sample A    Drug X             1         2 48.000000
#> 5 Sample A    Drug X            10         1  3.636364
#> 6 Sample A    Drug X            10         2  4.000000
str(expt)
#> 'data.frame':   80 obs. of  5 variables:
#> $ Sample       : Factor w/ 4 levels "Sample A","Sample B",..: 1 1 1 1 1 1 1 1 1 1 ...
#>  $ Treatment    : Factor w/ 2 levels "Drug X","Drug Y": 1 1 1 1 1 1 1 1 1 1 ...
#> $ Concentration: int  0 0 1 1 10 10 100 100 1000 1000 ...
#> $ Replicate    : int  1 2 1 2 1 2 1 2 1 2 ...
#> $ Data         : num  57 22 33 48 3.64 ...
```

The `Data` column contains the experimentally-measured data, which decreases with increasing concentration of drug.

```{r message=FALSE, warning=FALSE}
library(ggplot2)
ggplot(expt, aes(factor(Concentration ), log10(Data))) + 
   geom_point(
      aes(color = Sample),
      size = 3) + 
   facet_wrap( ~ Treatment) + 
   ylab("Response to drug treatment (log10 transformed)") + 
   xlab("Concentration of drug") +
   ggtitle("Samplewise response to drug treatment")

```

## Average replicates
(Claire)
include inter-replicate CV column in addition to avged values

## Normalize to control data

Once replicates are averaged, you often want to adjust the data relative to the control data. Examples of this kind of normalization include: 

* Background subtraction
* Percent change
* Fold change 
* ΔCt and ΔΔCt

In a typical data set, some rows will contain control data and some rows will contain experimental data. That's the case with the `expt` data set, where the rows with Concentration equal to 0 are vehicle controls (exposure to the liquid the drug was dissolved in, but without any drug). 

In this case, we want to view the response in each experimental row (the rows where Concentration isn't 0), as a percentage of the response in the vehicle control rows. There is a vehicle control for each combination of Sample and Treatment and we want to match them up correctly. 

The procedure we will follow is: 

1. Create a data frame with just the vehicle control rows.
2. Add the data from those rows as a new column to the `expt` data frame, matched by sample and treatment.
3. Calculate percent control response by operating on those two rows.

### Step 1: Create control data frame

!!!

Once the average replicates section is complete, norep.csv can be deleted and the read.csv line in the chunk below can be removed (since we'll use the `expt` generated above). Note that the `head(...)` output will need to be updated below.

!!!

```{r message=FALSE, warning=FALSE, results='hide'}
library(dplyr)
expt <- read.csv("norep.csv")

# subset to control rows
expt %>%
   filter(Concentration == 0) -> ctrl
head(ctrl)
#>     Sample Treatment Concentration Data
#> 1 Sample A    Drug X             0   16
#> 2 Sample A    Drug Y             0   89
#> 3 Sample B    Drug X             0   25
#> 4 Sample B    Drug Y             0   17
#> 5 Sample C    Drug X             0   27
#> 6 Sample C    Drug Y             0   75

# rename Data to ControlData and remove unneeded columns
ctrl %>%
   mutate(ControlData = Data) %>%
      select(-c(Concentration, Data)) -> ctrl
head(ctrl)
#>     Sample Treatment ControlData
#> 1 Sample A    Drug X          16
#> 2 Sample A    Drug Y          89
#> 3 Sample B    Drug X          25
#> 4 Sample B    Drug Y          17
#> 5 Sample C    Drug X          27
#> 6 Sample C    Drug Y          75
```

The `ctrl` data frame contains the information necessary to match the control data to the appropriate rows based on the values in the Sample and Treatment columns. 

### Step 2: Add control column to `expt`

```{r message=FALSE, warning=FALSE, results='hide'}
# merge expt and ctrl to add ControlData column matched by
# Sample and Treatment
expt <- merge(expt, ctrl)
head(expt)
#>    Sample Treatment Concentration        Data ControlData
#> 1 Sample A    Drug X             0 16.00000000          16
#> 2 Sample A    Drug X             1  5.50000000          16
#> 3 Sample A    Drug X            10  8.36363636          16
#> 4 Sample A    Drug X           100  0.65346535          16
#> 5 Sample A    Drug X          1000  0.06093906          16
#> 6 Sample A    Drug Y             0 89.00000000          89
```

As expected, the ControlData column has the same value for Sample A with Drug X at all concentrations, and this value is the measurement at a concentration of 0. For Sample A with Drug **Y**, there is a new ControlData value.   

### Step 3: Normalize

```{r message=FALSE, warning=FALSE, results='hide'}
# Add a new column containing the percent control response
expt <- mutate(expt, 
   PercentControl = Data / ControlData * 100)
head(expt)
#>     Sample Treatment Concentration        Data ControlData PercentControl
#> 1 Sample A    Drug X             0 16.00000000          16    100.0000000
#> 2 Sample A    Drug X             1  5.50000000          16     34.3750000
#> 3 Sample A    Drug X            10  8.36363636          16     52.2727273
#> 4 Sample A    Drug X           100  0.65346535          16      4.0841584
#> 5 Sample A    Drug X          1000  0.06093906          16      0.3808691
#> 6 Sample A    Drug Y             0 89.00000000          89    100.0000000
```

`expt` now contains a new column with the groupwise percent control response. As expected, PercentControl is always 100% where Concentration is 0. If desired, the control rows could now be removed with `expt <- filter(expt, Concentration != 0)`. 

Steps 1 and 2 would be very similar for any normalization procedure. In step 3, the equation would simply be changed to reflect the desired method. 

## Exclude outliers
Inter-replicate CV, too high, too low